Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2015-08-24 article.Rmd
Move critique back into text

The digital turn has made large datasets widely available, yet in the midst of abundance many humanists continue to deal with the problem of incompleteness and scarcity within their sources. Our method uses *abundant* existing datasets to infer additional information from *scarce* datasets. Its major contribution is to incorporate a historical method, one that takes into account how gendered naming practices have changed over time. This temporal approach provides much improved results over simpler, anachronistic, or ahistorical lookup methods. The gender package is, of course, far from perfect. By relying on historical datasets collected by state agencies, it reinforces a gender binary and larger power structure that obscures how people may identify themselves and how they experience the world. To borrow a phrase from Miriam Posner, the gender package does not necessarily meet the "radical, unrealized potential of digital humanities" [Posner 2015]. While we agree with this line of critique, we don't think it should deter researchers from using our method.^[While such critiques point out genuine problems, they also seldom contain concrete proposals that can be practically implemented in historically oriented research. Discussing the problem of gender/sex binaries in large corpora, Penelope Eckert argues that "binary sex-class gets us pretty far" and that "at the moment" she is pessimistic "whether there's any way we can code large corpora for something subtler about gender than sex-class" [Eckert 2012]. We thank Heather Froehlich for this citation.]

>>>>

The digital turn has made large datasets widely available, yet in the midst of abundance many humanists continue to deal with the problem of incompleteness and scarcity within their sources. Our method uses *abundant* existing datasets to infer additional information from *scarce* datasets. Its major contribution is to incorporate a historical method, one that takes into account how gendered naming practices have changed over time. This temporal approach provides much improved results over simpler, anachronistic, or ahistorical lookup methods. The gender package is, of course, far from perfect. By relying on historical datasets collected by state agencies, it reinforces a gender binary and larger power structure that obscures how people may identify themselves and how they experience the world. To borrow a phrase from Miriam Posner, the gender package does not necessarily meet the "radical, unrealized potential of digital humanities" [Posner 2015]. While we agree with this line of critique, we don't think it should deter researchers from using our method. While such critiques point out genuine problems, they also seldom contain concrete proposals that can be practically implemented in historically oriented research.^[Discussing the problem of gender/sex binaries in large corpora, Penelope Eckert argues that "binary sex-class gets us pretty far" and that "at the moment" she is pessimistic "whether there's any way we can code large corpora for something subtler about gender than sex-class" [Eckert 2012]. We thank Heather Froehlich for this citation.]
